Utilizing Parallel Workers: \\LLVM's Vectorization Plan

### 2024-11-22 paper/paper.tex
paper

LLVM's VPlan is currently being integrated into the existing Loop Vectorizer and is
already modeling most inner loop vectorizations and transformations~\cite{llvmvplanupdate}. 
Vectorization for outer loops is also in development and can be enabled by setting 
the \texttt{-enable-vplan-native-path} flag~\cite{llvmouterloop}. In the future, the plan will 
be to merge both of these loop vectorization paths into one. Merging the existing SLP Vectorizer
with VPlan is also planned.

@@

\subsection{Current State of VPlan}

In its current state (LLVM 19.1.4) VPlan is deeply integrated into LV. To better understand the
current structure of a VPlan, modelling a certain vectorization approach, we will use the
simple conditional loop from Figure~\ref{fig:inner-loop-vec} as an example.

@@

\subsubsection{General VPlan and Integration with LV}
The main effort of setting up the VPlan structure and integrating it first into LV has seen
tremendous effort over the last years. Florian Hahn presented future directions for 
VPlan~\cite{llvmvplanupdate}, which mostly evolve around transforming more LV-based decisions into
explicit VPlan-to-VPlan transformations. Additionally, a lot of cost model related decisions 
still end up being influenced by the old legacy LV cost model, however, transition to a VPlan-based 
cost model has already started.

@@

Outer loop vectorization support for VPlan is still experimental, seperated from the inner loop
LV integration, and runs on a divergent experimental path (\texttt{-enable-vplan-native-path}).
In the near future development on outer loop vectorization will continue and merging it with LV's
VPlan path is planned.

>>>>

LLVM's VPlan is currently being integrated into the existing Loop Vectorizer and is
already modeling most inner loop vectorizations and transformations~\cite{llvmvplanupdate}. 
Vectorization for outer loops is also in development and can be enabled by setting 
the \texttt{-enable-vplan-native-path} feature flag~\cite{llvmouterloop,llvmouterloopstatus}. 
In the future, the plan will be to merge both of these loop vectorization paths into one. 
Merging the existing SLP Vectorizer with VPlan is also planned.

@@

\subsection{State of VPlan (LLVM 19.1.4)}

In its current state VPlan is deeply integrated into LV.
When auto-vectorizing, LV starts by checking all the different loops in the IR.
If it finds one, it will try to vectorize it\\(\texttt{LoopVectorizePass::processLoop}).
Several factors might hinder vectorization for a given loop: 
(1) If the loop was already vectorized before, it won't be vectorized again. 
(2) The Legality phase checks if an how it is legal to vectorize the scalar
loop. (3) The loop's trip count might be too small to make vectorization profitable. (4) The cost
of runtime checks for e.g. pointer aliasing can be too high.

The initial VPlans are constructed based on the legality checks and the cost decisions the legacy
LV cost model made referencing the scalar IR\\(\texttt{LoopVectorizationPlanner::plan}). Then the
optimal Vectorization Factor (VF) and Interleave Count (IC) are picked. 
The best VPlan is picked based on the optimal VF. A VF of 1 means it is not
profitable to vectorize the loop, however, interleaving/unrolling can still end up modifying the
underlying IR.

The best VPlan is then executed\\(\texttt{LoopVectorizationPlanner::executePlan}). In case of
vectorization (VF $>1$), LV will also directly try to vectorize the epilogue loop.

\subsubsection{VPlan Structure}
To better understand the
underlying structure of a VPlan, modelling a certain vectorization approach, we will use the
simple conditional loop from Figure~\ref{fig:inner-loop-vec} as an example.

Figure~\ref{fig:inner-loop-scalar-ir} shows the IR of the main loop basic blocks \texttt{for.body},
\texttt{if.then}, and \texttt{for.inc}. Figure~\ref{fig:inner-loop-vplan-hcfg} shows the equivalent
hierarchical control-flow-graph (CFG) of a VPlan modelling a vectorization approach for this 
scalar loop.

@@

\subsubsection{VPlan and Integration with LV}
The main effort of setting up the VPlan structure and integrating it first into LV has seen
tremendous effort over the last years. Florian Hahn presented future directions for 
VPlan~\cite{llvmvplanupdate}, which mostly evolve around transforming more LV-based decisions into
explicit VPlan-to-VPlan transformations. Additionally, a lot of cost model related decisions 
still end up being influenced by the old legacy LV cost model, however, transition to a VPlan-based 
cost model has already started.

@@

Outer loop vectorization support via VPlan is still experimental, seperated from the inner loop
LV integration, and runs on a divergent experimental path (\texttt{-enable-vplan-native-path}).
In the near future development on outer loop vectorization will continue and merging it with LV's
VPlan path is planned.
